BATCAVE: Calling somatic mutations with a tumor- and site-specific prior

### 2020-01-13 ResponseToReviews.tex

\usepackage{xr}
\externaldocument{NAR_DRAFT}

\begin{document}
\vspace*{-0.5in}\hspace*{-0.2in}\includegraphics{LetterheadHeader}\vspace*{\baselineskip}

\noindent Dear Professor Herrero,

We appreciate the opportunity to revise our manuscript for publication in \emph{NAR Genomics and Bioinformatics}, and we thank you and the reviewers for the quality of the reviews. Below we respond in italics to the concerns of yourself and the reviewers. To aid re-review, we have included a manuscript with changes highlighted in the resubmission.

@@

\noindent Sincerely,\\
\includegraphics{signature}\\
Ryan Gutenkunst, PhD\\
Associate Professor and Associate Department Head\\
Department of Molecular and Cellular Biology\\
University of Arizona

\section*{Associate Editor}

Stressing on some of the points raised by the reviewers, I would suggest you include more details about the assumption and limitations of BATCAVE. In particular, the possibility of changes of mutational signatures over time can be quite striking in some cases, especially if the original tumour is driven by very specific mutagens like UV light or tobacco smoke. Ideally, the software should try to correct for these cases or at least warn the user about a possible shift in signatures/contexts.
\begin{response}
As we noted in the Discussion, BATCAVE does assume that the mutation profile is similar between mutations that are high-confidence (tend to be earlier) and low-confidence (tend to be later).
Modeling potential changes in the profile is beyond the scope of the current manuscript, but warning the user if such a shift exists is a great idea.
We have thus implemented a statistical test as to whether the mutation profile from the high-confidence mutations differs significantly from the low-confidence mutations.
If this test detects a difference, BATCAVE warns the user and outputs both profiles, so the user can assess the degree of difference and follow-up on any potential biological implications.

When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data (Fig.~S3).
Nevertheless, BATCAVE still performed extremely well (Table~1). 
\end{response}

Related to this, there is probably a minimum number of high-confidence mutations that are required by BATCAVE to produce reliable results. As with the previous point, ideally the batcaver should address or warn the user if the assumptions are violated.
\begin{response}
Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig.~3).
We have modified \texttt{batcaver} so that it now emits a warning if there are fewer than 500 high-confidence mutations in a sample.
\end{response}

@@

Lastly, as correctly pointed out by the second reviewer, the sequencing coverage only makes sense in the context of the purity of the samples and this information should be provided to the reader.
\begin{response}
The direct effect of contamination by normal cells on the BATCAVE algorithm is to bias the estimated allele frequencies that are inputs to Eq.~11 for estimating the overall mutation rate.
This bias can be simply corrected given an estimate of tumor purity.
We have thus added a \texttt{--tumor\_purity} option to \texttt{batcaver}.
When this is enabled, the estimated allele frequencies will be appropriately adjusted before applying Eq.~11.
See new text in Methods for details.

We also reanalyzed our real data examples, using purity estimates from the original publications (text and Table~S1).
Accounting for tumor purity changed the quantitative results slightly but did not change the qualitative conclusions.
We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.

@@

\end{response}

\section*{Reviewer 1}

Mannakee and Gutenkunst present BATCAVE, an algorithm that first estimates the tumor’s mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants based on mutational signatures. The concept behind this work is interesting. However, the difference in accuracy between MuTect and BATCAVE is relatively small except for the case when one has particularly high depth sequencing (500x), limiting the utility of the method.

Still, the 500x whole exome case does indicate that BATCAVE is better at identifying low frequency variants and thus can better reveal intratumoral heterogeneity, an important area of inquiry in cancer. Interestingly, the AUROC values for both MuTect and BATCAVE have room for improvement for the 500x whole exome case (Table 1). I suggest the authors further investigate how to improve the AUROC values. For example, in equation 4, it is unclear why the authors assume $\nu$ be uniformly distributed. Although this is also an assumption of MuTect, the BATCAVE approach inherently distinguishes mutations by allele frequency – since log-likelihood is closely associated to AF.  Perhaps a model in which $\nu$ follows a decreasing distribution vs. AF (e.g. in concordance with a model of a growing tumor) would yield better AUROC values.  
\begin{response}
A uniform prior over $\nu$ is appropriate when the goal is to make the best possible call at each potential variant site.
Conceptually, before examining the read data we have no information about the allele frequency at any individual site, hence a uniform prior.
Practically, a nonuniform prior would favor low-frequency and thus typically low-evidence sites, increasing the false positive rate.
If our goal were to best estimate the distribution of variant allele frequencies in the tumor, then a nonuniform $\nu$ would indeed be important.

@@

We discuss this issue in a new paragraph in Discussion.
\end{response}

In Rubanova et al (https://www.biorxiv.org/content/10.1101/260471v4.full), it is argued that there are changes in mutational signatures over time in cancers. While Mannakee and Gutenkunst mention this work, they should further comment on whether such changes in mutational signatures over time are small enough to be explained by the BATCAVER concept of re-evaluating mutations based on p-value.
\begin{response}
We cannot comment generally on the magnitude of mutation profile changes, because this is an developing area of research.
But to protect users, we have implemented and applied a new statistical test to detect changes in mutation profiles between high- and low-confidence mutations.
See our response to the Associate Editor for details.
\end{response}

@@

2. In case of multi-region sequencing data, used for studying spatial intra-tumor heterogeneity, usually the existing tools estimate very high levels of heterogeneity. I suppose this is because they are being very stringent in terms of calling mutations especially the ones with low vafs. In that case the people end up using other tools like bamreadcount etc. to check status of these mutations in the bam files and re-estimate the extent of heterogeneity. It will be important to see how BATCAVE performs in such datasets.
\begin{response}
We agree that multi-region data are increasingly important.
In fact, the breast cancer data we test on come from a multi-region study. 
Because we built BATCAVE on top of the single-sample caller MuTect, we called variants independently across the samples in that study, and BATCAVE performed well.
Unfortunately, the sequencing in that study is not deep enough to rigorously evaluate how estimates of heterogeneity would change with and without BATCAVE.

More broadly, the BATCAVE algorithm could be implemented in a multi-regional caller that takes explicit advantage of the relationship between samples.
But such an implementation is beyond the scope of the current manuscript.

We have added a new paragraph in Discussion regarding multi-region samples, to address these issues.
\end{response}

3. Do the mutation signatures (example smoking signatures in lung cancer or UV signatures in melanoma) still appear in the more stringent BATCAVE set of mutations?
\begin{response}
In response to this and other review comments, we have added a statistical test as to whether the mutation profile changes between high- and low-confidence sets of mutations.
And in the AML data, we do see such a change (Fig.~S3).

Approaches such as deconstructSigs can decompose an individual sample's mutational profile into a set of mutation signatures corresponding to different processes.\footnote{Rosenthal et al. (2016) \emph{Genome Biology} 17:31}
In practice, however, we have found these approaches to be relatively unstable.
Therefore, we do not believe it will be illuminating to apply them in the present manuscript.
\end{response}

### 2020-01-22 Mannakee_zip/NAR_DRAFT.tex
Clean up for final files to NAR Genomics and Bioinformatics

\usepackage{xspace}
\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
\newcommand{\batcave}{BATCAVE\xspace}
\newcommand{\beginsupplement}{%
        \clearpage
        \onecolumn
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Enter dates of publication
\copyrightyear{x}
\pubdate{x}
\pubyear{x}
\jvolume{x}
\jissue{x}

@@

\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}

@@

\begin{abstract}
Detecting somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
Mutations at low allelic frequency, those present in only a small portion of tumor cells, are particularly difficult to detect.
Many algorithms have been developed to detect such mutations, but none models a key aspect of tumor biology.
Namely, every tumor has its own profile of mutation types that it tends to generate.
We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
We also present an R implementation of the algorithm, built on the popular caller MuTect.
Using simulations, we show that adding the \batcave algorithm to MuTect improves variant detection.
It also improves the calibration of posterior probabilities, enabling more principled tradeoff between precision and recall.
We also show that \batcave performs well on real data.
Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
More broadly, the algorithm can be added to other variant callers, and it can be extended to include additional biological features that affect mutation generation.
\end{abstract}

@@

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our \batcave algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

\begin{figure}
\centering
  \includegraphics{figures/real_signatures_only.pdf}
  \caption{Real tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

@@

Assuming that the identity of the alternate allele and its allele frequency are independent and that $\nu$ is uniformly distributed, Eq.~\ref{eqn:1} becomes
\begin{equation}  \label{eqn:2}
  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
\end{equation}
The focus of \batcave is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

@@

We model the observed mutation profile $\mathrm{S}$ as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.

@@

The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
The R implementation of \batcave ships with pre-computed tables for both human whole exomes and whole genomes.

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p_{ur}$.
All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ and $p_{ur}$ are free parameters.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
For our simulated data analyses, we set $p_{ur}=1$, and for our real data analyses, we set $p_{ur}$ to the estimated purity.

@@

\subsection{Likelihood function}
The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq. \ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all analyses in this paper, so we have
\begin{equation}
  \label{eqn:tlod}
    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
\end{equation}
The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
\begin{equation}
  \label{eqn:computed_posterior}
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
\end{equation}
Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
\begin{equation}  \label{eqn:mutect_posterior}
  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
\end{equation}

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package called \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

An assumption of the \batcave algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
If the two profiles differ at a significance threshold of $\alpha=0.05$, \texttt{batcaver} issues a warning and outputs the profiles.
%The user can then assess them and follow-up on any biologically interesting differences.

\begin{figure}
\centering
  \includegraphics{figures/sim_signatures_only.pdf}
  \caption{Simulated tumor mutation profiles.
As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.\\
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

>>>>

% Enter dates of publication
\copyrightyear{x}
\pubdate{x}
\pubyear{x}
\jvolume{x}
\jissue{x}

@@

\title{BATCAVE: Calling somatic mutations with a tumor- and site-specific prior}

@@

\begin{abstract}
Detecting somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
Mutations at low allelic frequency, those present in only a small portion of tumor cells, are particularly difficult to detect.
Many algorithms have been developed to detect such mutations, but none models a key aspect of tumor biology.
Namely, every tumor has its own profile of mutation types that it tends to generate.
We present BATCAVE (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the individual tumor mutational profile and mutation rate then uses them in a prior for evaluating potential mutations.
We also present an R implementation of the algorithm, built on the popular caller MuTect.
Using simulations, we show that adding the BATCAVE algorithm to MuTect improves variant detection.
It also improves the calibration of posterior probabilities, enabling more principled tradeoff between precision and recall.
We also show that BATCAVE performs well on real data.
Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
More broadly, the algorithm can be added to other variant callers, and it can be extended to include additional biological features that affect mutation generation.
\end{abstract}

@@

Here we present an enhanced variant-calling algorithm that uses the biology of each individual tumor's mutation profile to improve identification of low allelic frequency mutations.
Our BATCAVE algorithm first estimates the tumor's mutation profile and mutation rate using high-confidence variants and then uses them as a prior when calling other variants.
Our R implementation of the algorithm, \texttt{batcaver}, takes output from the MuTect variant caller as input and returns the posterior probability that a site is variant for every site observed by MuTect.
Using both simulated and real data, we show that the addition of a mutation profile prior to MuTect produces a superior variant caller.
Our algorithm is simple and computationally inexpensive, and it can be integrated into numerous other variant callers.
Broad adoption of our approach will enable more confident study of low allelic frequency mutations in tumors in both research and clinical settings.

\begin{figure}
\centering
  \includegraphics{real_signatures_only.pdf}
  \caption{Real tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

@@

Assuming that the identity of the alternate allele and its allele frequency are independent and that $\nu$ is uniformly distributed, Eq.~\ref{eqn:1} becomes
\begin{equation}  \label{eqn:2}
  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
\end{equation}
The focus of BATCAVE is to provide a tumor- and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

@@

We model the observed mutation profile $\mathrm{S}$ as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In BATCAVE we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely to any allele and in any context.

@@

The prior probability of each particular trinucleotide context $p(C = c)$ is computed simply as the proportion of sequenced trinucleotide contexts that have context $c$.
The R implementation of BATCAVE ships with pre-computed tables for both human whole exomes and whole genomes.

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
Sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p_{ur}$.
All estimated allele frequencies are then multiplied by a factor of $1/p_{ur}$ before being used in Eq.~\ref{eqn:mut_rate}.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of BATCAVE, $f_{min}$ and $p_{ur}$ are free parameters.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
For our simulated data analyses, we set $p_{ur}=1$, and for our real data analyses, we set $p_{ur}$ to the estimated purity.

@@

\subsection{Likelihood function}
The current implementation of BATCAVE builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq. \ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all analyses in this paper, so we have
\begin{equation}
  \label{eqn:tlod}
    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
\end{equation}
The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
\begin{equation}
  \label{eqn:computed_posterior}
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
\end{equation}
Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
\begin{equation}  \label{eqn:mutect_posterior}
  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
\end{equation}

\subsection{Implementation}
We have implemented the BATCAVE algorithm as an R package called \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

An assumption of the BATCAVE algorithm is that the mutation profile of low-confidence mutations is similar to that of the high-confidence mutations from which the prior is inferred.
To test this assumption, \texttt{batcaver} performs a statistical hypothesis test comparing these two mutation profiles \cite{Plunkett2019}.
If the two profiles differ at a significance threshold of $\alpha=0.05$, \texttt{batcaver} issues a warning and outputs the profiles.
%The user can then assess them and follow-up on any biologically interesting differences.

\begin{figure}
\centering
  \includegraphics{sim_signatures_only.pdf}
  \caption{Simulated tumor mutation profiles.
As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.\\
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to BATCAVE for prior estimation and rescoring.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and BATCAVE, we used the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and BATCAVE.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we analyzed the three biological replicates from each of six tumors (Table~\ref{table:01}), running BATCAVE separately on each of the eighteen samples, and we combined results to generate precision-recall curves.
Shi et al.\ estimated sample purities of between 26\% and 80\%, and we used these estimates in \texttt{batcaver}.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCh37, so we used that reference.

\section{RESULTS}
We implemented BATCAVE as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
BATCAVE extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The BATCAVE algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.
